Analysis of Toronto Homeless Shelter System:

### 2023-02-03 outputs/paper/paper.qmd
Final touches

---
title: "Analysis of Toronto Homeless Shelter System:"
subtitle: "A closer look at hidden homelessness"
author: "Huda Sahaf"
thanks: "Code and data are available at:" 
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
bibliography: references.bib
---

@@

For people experiencing homelessness in the city of Toronto, there are several options for help, most of which are funded by the government. These services are of many different types: homeless shelters, drop-in programs to access the necessities, 24-hour respites, warming spaces, and programs to get help with housing.Despite these programs, in 2021, there were 18,096 people who were experiencing homelessness in Toronto [@homie]. Homelessness is a serious problem, and unfortunately homeless shelters as well as most of the services that are offered by the city of Toronto are not a sustainable solution. The causes for homelessness are many and these service options do not address them.

These services are made up of a network of 53 sites that offer services that are approved by the city of Toronto[@numsites]. Some of these sites are run and managed by the city itself, and others are community sites that are provided for by the city. Shelters offer a possible referral to other community services based on need, counselors to help find permanent housing, and mental and physical health support services. 24-hour respite sites are also similar in that they also offer referrals for other services based on the need of the individual.

@@

 To conduct the analysis on the city of Toronto’s homeless shelter system data, R [@citeR] will be used. The data set itself was downloaded from `opendatatoronto` [@odt]. The packages `tidyverse` [@tidyv], `dplyr` [@dan], and `janitor` [@janice] will be used to read in, clean and manipulate the data. Any graphs and tables displayed were created using `ggplot2` [@greg], `knitr` [@nat], `kableExtra` [@konrad], and coloring options were made possible using the `RColorBrewer` [@ryan] package.

@@

**Chronic:** This category refers to people who are experiencing chronic homelessness. More specifically, chronic homelessness is defined as someone who has been reported to have 180 overnight stays in any or multiple of the homeless service facilities in the last year, or, in the past five years, the person has a total number of reported stays of 546 nights.

**Families:** These are reported individuals who are staying at overnight facilities appointed by their families

@@

There are also age breakdowns, gender demographics and population percentage available within this data set.

# Discussion

The purpose of the data is to provide a holistic view of the homeless shelter system in the city of Toronto. This data set specifically documents the flow of people in the homeless shelter system due to the fact that people are constantly coming in and out of the system. There are many different types of services, and as such SMIS reflects the entry and exit of an individual in the system on a monthly basis.

According to the city of Toronto, this data is used to help measure progress towards one of the goals of the government: to make sure that homelessness in this city becomes “rare, brief, and non-recurring [@rare].” There are also different types of services that individuals often move between, which is why the numbers in the Shelter Management Information System data are reported to be higher.

@@

```{r}
all_newly_id <- 
  all_newly_id |>
  left_join(newly_identified_18)
```

@@

```{r}
#| echo: false
#| include: false
#| label: active_homeless_22
#| fig-cap: Actively Homeless People 2022
cleaned_shelter_data|>
  select(population_group, Year, Month, actively_homeless)|>
  filter(population_group == "All Population")|>
  group_by(Year)|>
  ggplot(mapping = aes(x = Month, y = actively_homeless, fill = Year, label = actively_homeless)) + 
  geom_bar(stat = "identity") +
  theme_light() +
  geom_text(size = 2, position = position_stack(vjust = 0.6)) +
  labs(
    title = "Number of Actively Homeless Individuals by Month",
    x = "Month",
    y = "Number of Actively Homeless"
  )
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| tbl-cap: "wow bro."
#| label: tbl-testing 
#| fig-width: 8
#| fig-height: 4
#Actively Homeless Table
cleaned_shelter_data|>
  select(population_group, Year, actively_homeless)|>
  filter(population_group == "All Population")|>
  group_by(Year)|>
  summarise(average = mean(actively_homeless))|>
  knitr::kable()
```

@@

Also bills and their average (@tbl-testing). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work.)

@@

There are limitations with regards to this data set that must be addressed. Firstly, the way that the city of Toronto collects shelter data is very reflective of “point-in-time counts. [@lim]” The problem with this type of data collection is that it lacks context. It is relevant only in that it reflects what the homeless shelter system looks like at any given point in time. They do not fully account for hidden homelessness, and this impacts the numbers reported, especially in women and youth.

Another limitation is in the reporting of Indigenous individuals in the data. There are many columns in which the row belonging to the Indigenous population group simply does not exist. Due to this, in Figure 1, there is no column for Indigenous data. The city of Toronto has not acknowledged this as an issue, and as a result, there is no reason given for why this inconsistency exists. It is not just for the newly identified category that this is a problem, but for all situation data categories.

A characteristic of this dataset is that the population group definitions are all extremely general. There is no defined meaning to the category “Non-Refugees”, and whether or not there is room for overlap in the categories “Non-Refugees” and “Single Adult”. There is very little detail given as to the demographic break down of the population groups, they seem situationally narrow and do not give an idea of the identity’s of the people themselves.

@@

\appendix

# Appendix {.unnumbered}

# Additional details

>>>>

---
title: "Analysis of Toronto Homeless Shelter System:"
subtitle: "A closer look at hidden homelessness"
author: "Huda Sahaf"
thanks: "Code and data are available at:" 
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "The homeless population in Toronto is rapidly increasing at a rate the homeless shelter system is ill-quipped for. The aim of this investigation to use the city of Toronto’s Shelter Management Information System data to analyze any changes to the number of newly identified homeless people, to take a closer look at the population groups that remain unaccounted for in the data. Understanding how hidden homelessness remains unseen in the data will help make sense of the impact on the future of homeless individuals in Toronto."
format: pdf
bibliography: references.bib
---

@@

For people experiencing homelessness in the city of Toronto, there are several options for help, many of which are funded by the government. These services are of several different types: homeless shelters, drop-in programs to access necessities, 24-hour respites, warming spaces, and programs to get help with housing. Despite these programs, in 2021, there were 18,096 people who were experiencing homelessness in Toronto [@homie]. Homelessness is a serious problem, and unfortunately homeless shelters as well as most of the services that are offered by the city of Toronto are not a sustainable solution. The root causes for homelessness are many and these service options do not address them.

These services are made up of a network of 53 sites that offer services approved by the city of Toronto[@numsites]. Some of these sites are run and managed by the city itself, and others are community sites that are provided for by the city. Shelters offer a possible referral to other community services based on need, counselors to help find permanent housing, and mental and physical health support services. 24-hour respite sites are also similar in that they also offer referrals for other services based on the need of the individual.

@@

To conduct the analysis on the city of Toronto's homeless shelter system data, R [@citeR] will be used. The data set itself was downloaded from `opendatatoronto` [@odt]. The packages `tidyverse` [@tidyv], `dplyr` [@dan], and `janitor` [@janice] will be used to read in, clean and manipulate the data. Any graphs and tables displayed were created using `ggplot2` [@greg], `knitr` [@nat], `kableExtra` [@konrad], and coloring options were made possible using the `RColorBrewer` [@ryan] package.

@@

**Chronic:** This category refers to people who are experiencing chronic homelessness, which is defined as people who have been reported to have 180 overnight stays in any or multiple of the homeless service facilities in the last year.

**Families:** These are reported individuals who are staying at overnight facilities appointed by their families.

@@

# Discussion
The purpose of the data is to provide a holistic view of the homeless shelter system in the city of Toronto. This data set specifically documents the flow of people in the homeless shelter system due to the fact that people are constantly coming in and out of the system. There are many different types of services, and as such SMIS reflects the entry and exit of an individual in the system on a monthly basis.

According to the city of Toronto, this data is used to help measure progress towards one of the goals of the government: to make sure that homelessness in this city becomes "rare, brief, and non-recurring [@rare]." There are also different types of services that individuals often move between, which is why the numbers in the Shelter Management Information System data are reported to be higher.

@@

```{r}
#| echo: false
#| include: false
# 2018
all_newly_id <- newly_identified_18 |>
  left_join(newly_identified_18)

# 2020
all_newly_id <- all_newly_id|>
   left_join(newly_identified_20)
# 2022
all_newly_id <- all_newly_id|>
  left_join(newly_identified_22)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| tbl-cap: "Mean Number of Newly Identified Homeless People per Year."
#| label: tbl-new
all_newly_id|>
  knitr::kable()
```

@@

```{r}
#| echo: false
#| label: fig-active_homeless
#| fig-cap: Total Actively Homeless People 2018-2022

cleaned_shelter_data$Month <- factor(cleaned_shelter_data$Month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

cleaned_shelter_data|>
  select(population_group, Year, Month, actively_homeless)|>
  filter(population_group == "All Population") |>
  group_by(Year)|>
  ggplot(mapping = aes(x = Month, y = actively_homeless, fill = Year, label = actively_homeless)) +
  geom_bar(stat = "identity") +
  theme_light() +
  geom_text(size = 2, position = position_stack(vjust = 0.5)) +
  labs(
    title = "Number of Actively Homeless Individuals by Month",
    x = "Month", 
    y = "Number of Actively Homeless"
  ) +
  scale_fill_brewer("Set3")

```{r}
#| message: false
#| warning: false
#| echo: false
#| tbl-cap: "Mean Number of Homeless People per Year."
#| label: tbl-testing 
#Actively Homeless Table
cleaned_shelter_data|>
  select(population_group, Year, actively_homeless)|>
  filter(population_group == "All Population")|>
  group_by(Year)|>
  summarise(average = mean(actively_homeless))|>
  knitr::kable()
```

@@

While the categorization of the population groups may not tell us much in terms of specific groups that are at risk of being homeless, the total numbers are not fluctuating drastically over time. (@tbl-testing) is a table of the mean number of actively homeless people, over the past five years. (@tbl-new) is The numbers decreased in 2019 and 2020 but are now back up to what the number was in 2019. [@fig-active_homeless] graphs the number of total actively homeless people in all population categories by month over the course of five years. As shown in [@fig-active_homeless], December is the month in which the total number of actively homeless people rises every year, and it has been the highest in December of last year. However, these reports are not consistent with the reality of thousands of people in Toronto. Increasing rent and inflation has forced many more people on the streets.

It is not reflected in the data surrounding newly identified homeless individuals nor actively homeless people, despite the gravity of the situation. It may be a limitation in the data in that people who are turned away from the shelter system are not included in the reports, but their presence is drastically increasing. According to the Toronto Star, the homeless shelter system within this city is currently unable to support even half of the homeless population, despite the fact that the numbers are rapidly growing. [@star] Getting permanent housing is also a very difficult process since there is a 10-12 year waitlist on social housing. [@star] This means that the lack of correct reporting has an impact that is far reaching. A lack of a reported increase in newly identified homeless individuals results in no changes made by the city of Toronto to improve the social housing system, which means more people will be remaining actively homeless, and none of these people are accounted for in the data.

@@

There are limitations with regards to this data set that must be addressed. Firstly, the way that the city of Toronto collects shelter data is very reflective of "point-in-time counts. [@lim]" The problem with this type of data collection is that it lacks context. It is relevant only in that it reflects what the homeless shelter system looks like at any given point in time. They do not fully account for hidden homelessness, and this impacts the numbers reported, especially in women and youth.

Another limitation is in the reporting of Indigenous individuals in the data. There are many columns in which the row belonging to the Indigenous population group simply does not exist. Due to this, in Figure 1, there is no column for Indigenous data. The city of Toronto has not acknowledged this as an issue, and as a result, there is no reason given for why this inconsistency exists. It is not just for the newly identified category that this is a problem, but for all situational data categories.

A characteristic of this data set is that the population group definitions are all extremely general. There is no defined meaning to the category "Non-Refugees", and whether or not there is room for overlap in the categories "Non-Refugees" and "Single Adult". There is very little detail given as to the demographic break down of the population groups, they seem situationally narrow and do not give an idea of the identities of the people themselves.

@@
